### 2016-01-06 ArticuloCSDyCM/Paper01.tex
uura papaer

The concepts presented here require first obtaining the CSD of the recorded LFP. Any method would conceptualy present the same utility, but thanks to the density and scales of the BioCAM 4096, in our case a ``classical'' numerical derivation of the CSD is the most adecuate route to take. A numerical Laplacian Operator which reduces the cross efect of rectangular grids is the convolution of the data with the following matrix \cite{Lindberg90}:
\begin{equation}
\nabla^2_{1/3}=(2/3)
\begin{pmatrix}
  0 & 1 & 0 \\
  1 & -4 & 1 \\
  0 & 1 & 0
\end{pmatrix}
+1/3
\begin{pmatrix}
  0.5 & 0 & 0.5 \\
  0 & -2 & 0 \\
  0.5 & 0 & 0.5
\end{pmatrix}  
\end{equation}
At the border we duplicate the last rows of data and take the results knowingly of the border effects. As this sort of operators are extremelly sensitive to noice, a Gaussian blur is applied before presenting the graphical data. This softening of the edges in the CSD image also represents our ignorance of the finer details of the components of the distribution (see section bla bla).

More sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptios at the scales involved in these data, and turn out to be computer-costly for high density MEAs and dense temporal sampling.

@@

One of the most obvious uses of the CSD in place of the LFP has been overlocked, and this might have to do that samplings at lesser spatial density did not permit extending the implications of the concept of density, namely, that one of ``masses'', or, in this case, charges. Center of Mass Analysis (CMA) has been previously been used in order to track putative ``trayectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Manjarrez07}. In such works the choosen ``mass'' was an heuristic measure of activity adecuate for representing the corse-grain picture of displacement of activity or some unit of ``active information''. But these present the usual problem of justifing the choice maide on practical grounds. For calculating a vectorial average, the adecuate choice of wheights per vector is a density, a positive quantity that measures the amount of something per unit volume. Current Source Density, as it name implies, is the density of sinks and sources of charged current per unit volume. CSD sepparates naturally the measured activity into two distinct sets, namely, the set of all sources and the set of all sinks, separated by the zero set. This is in contrast to the LFP, where the zero value of the function has no solid meaning as electrical potentials can be measured to arbitrary ground values.

A visual inspection of the CSD colormap at a given instant in time shows that both sources and sinks appear in rather large and significant patches. These appear to correspond to the expected double and triple poles of fyring neurons in structured tissue \cite{} and inspired us to sepparate the CSD subsets into their \emph{disjoint components}. This concept is used in mathematics to denote subsets of a geometric defined set that do not touch each other. Their rough correspondence to poles of fyring neurons would be most usefull in tracking the succesive locus of the activity.

To clarify the concept and how we are we using it an example is in order. In figure \ref{ejemplodisjuntos} we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set A consists of a single component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set B is made of two disjoint components, labeled B1 and B2. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one componet, then again we can trace a path joining them in the set. So the set B consists of two disjoint components, but each component can be regarded as a connected set on their own. Likewise the set C is made of three disjoint components.

CSD divides the set of measuring points into two subsets: the set of all sources and the set of all sinks, i.e, the set of all measuring points that have a CSD positive value, and the set of those who have a negative CSD value. The zero set is a border between them, but due to the finite precision of the measurement (and the validity of the mean field approach!) these sets can make contact with each other. Still, noice between patches corresponding to a single set could make a ``bridge'' between components, making identification of poles inexact. In order to avoid this we create a ``fat zero'' set of all the recording points whose CSD values are inside a small error interval around zero. This sepperates even more the components of the set, and helps us to simplify the following analysis.

@@

Densities provide the adecuate conceptual tool to obtain ``Centers of Mass'' or vector averages. A density is a measure of a quantity per spatial spread, at precise locus in space. CSD has, in contast to mass density, negative values. But these signs are a matter of convention, it is more important to say that it has two different cualitative propiertes and that sepparates the domain into two sets. Each set has then a total ``mass'', the integral of the absolute value of the CSD. Then the concept of ``Center of Mass'' applies naturally. Once that we sepparate the two sets and apply the concept in each one independently, even the sign convention of the densities seases to be a problem, as it cancels in the formula for the CM:
\begin{equation}
\langle x(t) \rangle =\frac{\sum_j x_{j} (t) I_{j} (x,t)} {\sum_j I_{j}(x,t)}
\end{equation}
An instantaneous Center of Mass, thus calculated, would present little if none physiological interpretation. The activity could be spread over non-convex sets (which would include disjoint and concave sets) and so the vector average, the Center of Mass, would be located in points where there is no activity. In our example the use of the CA complex of the hippocampus proves an ideal example. This structure is concave, and during epileptic burst, can have activity spread over many disjoint sites simultaneously. We must not take out of sight the phenomena which is producing the CSD distribution: neurons have a finite size, and the effect of their action a finite extent. Every patch that we can delimit as a disjoint component of the Sources/Sink sets indicates a coherent, joint active group of units. A center of mass for each disjoint component would then have much

>>>>

The concepts presented here require first obtaining the CSD of the recorded LFP. Any method would present the same utility. Thanks to the density and scales of the BioCAM 4096, in our case a ``classical'' numerical derivation of the CSD is the most adecuate route to take. A numerical Laplacian Operator which reduces the cross efect of rectangular grids is the convolution of the data with the following matrix \cite{Lindberg90}:
\begin{equation}
\nabla^2_{1/3}=(2/3)
\begin{pmatrix}
  0 & 1 & 0 \\
  1 & -4 & 1 \\
  0 & 1 & 0
\end{pmatrix}
+1/3
\begin{pmatrix}
  0.5 & 0 & 0.5 \\
  0 & -2 & 0 \\
  0.5 & 0 & 0.5
\end{pmatrix}  
\end{equation}
At the border we duplicate the last rows of data and take the results knowingly of the border effects. As this sort of operators are extremely sensitive to noise, a Gaussian blur is applied before presenting the graphical data. This softening of the edges in the CSD image also represents our ignorance of the finer details of the components of the distribution (see section bla bla).

More sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptions at the scales involved in our data, and turn out to be computer-costly for high density MEAs and dense temporal sampling.

@@

One of the most obvious uses of the CSD in place of the LFP has been overlooked. This might have occurred because  samplings at lesser spatial density did not permit extending the implications of the concept of density. Density is \emph{quantifiable entity per unit of space}, or, in other words, concentration of something distributed in space. In this case, the ``something'' is the sources and sinks of electric current, which can be identified with the interchange of charged ions from the inside to the outside of cellular membranes. As these are indicators of activity in the neurons, the CSD is also, indirectly, a \emph{``density of local activity''}. The advantages of this representation of the activity in place of the LFP have been stated many times (better spatial resolution, mostly), but is use as \emph{density} has been overlooked.

Center of Mass Analysis (CMA) has previously been used in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Manjarrez07}. In such works the chosen ``mass'' was an heuristic measure of activity adecuate for representing the corse-grain picture of displacement of activity or some unit of ``active information''. But these present the usual problem of justifing the choice maide on practical grounds. For calculating a vectorial average, the adecuate choice of wheights per vector is a density, a positive quantity that measures the amount of something per unit volume. CSD sepparates naturally the measured activity into two distinct sets, namely, the set of all sources and the set of all sinks, separated by the zero set. The sepparation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distintion between them is not. 
This is also in contrast to LFP, where the zero of the function has no precise meaning as electrical potentials can be measured from arbitrary ground values.

A visual inspection of the CSD colormap at a given instant in time shows that both sources and sinks appear in rather large and significant patches. These appear to correspond to the expected double and triple poles of fyring neurons in structured tissue \cite{} and inspired us to separate the CSD subsets into their \emph{disjoint components}. This concept is used in mathematics to denote subsets of a given geometric set that do not touch each other. Their rough correspondence to poles of firing neurons would be most useful in tracking the successive locus of the activity.

To clarify the concept and how we are we using it an example is in order. In figure \ref{ejemplodisjuntos} we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set A consists of a single component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set B is made of two disjoint components, labeled B1 and B2. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set B consists of two disjoint components, but each component can be regarded as a connected set on their own. Likewise the set C is made of three disjoint components.

CSD divides the set of measuring points into two subsets: the set of all sources and the set of all sinks, i.e, the set of all measuring points that have a CSD positive value, and the set of those who have a negative CSD value. The zero set is a border between them, but due to the finite precision of the measurement (and the validity of the mean field approach!) these sets can make contact with each other. Still, noise between patches corresponding to a single set could make a ``bridge'' between components, making identification of poles inexact. In order to avoid this we create a ``thick zero'' set of all the recording points whose CSD values are inside a small error interval around zero. This sepperates even more the components of the set, and helps us to simplify the following analysis.

@@

Densities provide the adequate conceptual tool to obtain ``Centers of Mass'' or vector averages. A density is a measure of a quantity per spatial spread, at precise locus in space. CSD has, in contast to mass density, negative values. But these signs are a matter of convention, it is more important to say that it has two different cualitative propiertes and that sepparates the domain into two sets. Each set has then a total ``mass'', the integral of the absolute value of the CSD. Then the concept of ``Center of Mass'' applies naturally. Once that we sepparate the two sets and apply the concept in each one independently, even the sign convention of the densities seases to be a problem, as it cancels in the formula for the CM:
\begin{equation}
\langle x(t) \rangle =\frac{\sum_j x_{j} (t) I_{j} (x,t)} {\sum_j I_{j}(x,t)}
\end{equation}
An instantaneous Center of Mass, thus calculated, would present little or none physiological interpretation. The activity could be spread over non-convex sets (which would include disjoint and concave sets) and so the vector average, the Center of Mass, would be located in points where there is no activity. In our example the use of the CA complex of the hippo campus proves an ideal example. This structure is concave, and during epileptic burst, can have activity spread over many disjoint sites simultaneously. We must not take out of sight the phenomena which is producing the CSD distribution: neurons have a finite size, and the effect of their action a finite extent. Every patch that we can delimit as a disjoint component of the Sources/Sink sets indicates a coherent, joint active group of units. A center of mass for each disjoint component would then indicate an average locus for this active group, a putative center of activity to which we can give a position, intensity and time. An instantaneous snapshot of such information would depict the centers of activity of an ordered structure of neurons and their relative intensities. But it is in the successive analysis of a series of such snapshots that this analysis can show its true power. The apparent , movement, and disappearance of such centers could reveal very subtle details of the dynamics of neural activity. Very small, fast displacements of the Centers of mass, could reveal slight asynchronism between neurons which could belong to the same active group. Large, slower than a few firing periods, could indicate physiological connections between various groups.

\section{The Recipe}

In order to perform the analysis described here, we summarize the steps.
\begin{enumerate}
\item Acquire high density electrophysiological data. The data must be two or three dimensional in space. Distance between the recording sites must be of the order of the size of the typical neurons involved, and such sites must be on a dense enough grid to perform differential operation over them. The sampling frequency should be enough to detect subthreshold activity, of at least 5kHz. Also, the data should come from structured tissue. Unordered, highly homogeneous tissue, will not yield significant results. 
\item Obtain, frame by frame, the CSD from the data. Any method may be used, but computer expensive techniques are discouraged if the data is dense enough. Common sense and tests are guidelines here.
\item Separate the CSD data into two three sets: sources, sinks, and border or indiscernible (below error) data.
\item Perform disjoint component analysis in the sources and sinks sets, frame by frame.
\item  Obtain the CM for each disjoint component.
\item For each frame and its successor, and for each CM, perform a ``most probable successor'' detection. Here we must put numerical criteria according to the data avaible, i.e. how far awat and how different in intensity should be allowed a CM to be at time $t+1$ to be considered succesor in time of another at time $t$.
  \item plot the data and try to figure what it means.
\end{enumerate}

A series of experiments done in the laboratory on rat's hippocampus provided us with the data that we shall use to illustrate the CSD and CM analysis proposed above.

\section{Application to epileptic-like activity in the hippocampus}
Chronologically speaking, we performed first the analysis to the most complicated data (numerically and qualitatively), which where a series of epileptic burst that occured after 15 minutes of application of 4AP to living slices of hippocampus. The burst occured with activity spreading over many sites simultaneously. This turned out to be a test of the applicability and robustness of our method. The necessity of having good error value for the ``thick zero'' set was more apparent here than anywhere else. Also the posibilty to create new structures of numerical data was seen here.  But also the most strickig features of the propagation of activity, coded now as trajectories of the CM, could be seen here, as the epileptic burst covered the whole CA structure.

\section{Application to evocated activity in idem.}

The evocated activity seem to yield much cleaner results, as we can now compare the trajectories of successive, very similar events. Our CSD and CM analysis can be used to detect consistent features in these experiments, such as the direction the waxing and wanning of activity. Small deviations from the trajectory in successive experiments that are above the numerical error could indicate inconsistencies in the network below.

\section{Application to spontaneous activity in idem.}

We had one experiment in which we could trace very nicely the data on DG and CA simultaneously without adding excitatory drugs.
